Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2011-05-26 Text/migration.tex
Added URL
	modified:   Text/migration.tex

git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@83 3b7f7548-5c66-0410-83ab-3997167635fc

\begin{abstract}
Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that undergoes gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially close to marginally instability. In one simulation, the planet is inserted prior to the eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases with embedded planets, there are times when the planet's radial motion is slow and varies in direction. At other times, when the planet appears to be interacting with strong spiral modes, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. Migration in both cases appears to stall near the inner Lindblad resonance of a dominant low-order mode. Planet orbit eccentricities fluctuate rapidly between about 0.02 to 0.1 throughout the GI-active phases of the simulations. 
\end{abstract}

@@

{\it The $t = 0$ Planet Run.} Figure \ref{fig:Am} shows that inclusion of even this modest mass planet ($0.7 \%$ of $M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4 ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum at $26$ AU. Modes of cos($m\phi$) symmetry with $m =$ 4 and 5 dominate. An $m = 3$ (three-armed) spiral also grows, but somewhat more slowly, and lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake within about an ORP in which $m = 3$ is a dominant component ($\sim 5$ \% global density perturbation). This nonlinear seeding causes $m = 3$ to dominate the GI burst, which also occurs $\sim$ 2 ORP earlier. Because of our initial placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly close to that of the planet's orbit period. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP, planet migration is significantly affected. Figure \ref{fig:a} shows the evolution of the planet's radial position. From 0 to 2.5 ORP, the planet is torqued primarily by its own wake and migrates inward. 
Beginning at about 3 ORP, the planet interacts with the now nonlinear $m = 3$ GI mode. At first, the planet gains angular momentum and moves outward, but, from 4 to 8 ORP, a time interval of  about 720 yr, the plant experiences a negative torque and plunges from 23 to 17 AU.

After $t =$ 8 ORP, the main burst is over, and the disk transitions into its asymptotic state, where modes of many $m$-values become comparably strong (see Fig. \ref{fig:Am}). The planet's radial migration apparently stalls at about 16 to 17 AU. From an analysis of periodicities present in the gas disk between 8 and 12 ORP, the planet lies a few AU inside the inner Lindblad resonance of a strong $m = 2$ mode with CR at 29 AU in this early part of the asymptotic phase.

{\it The $t = 10$ Planet Run.} The motion of the planet in this case is more difficult to interpret. We see intervals of fairly rapid inward or outward migration over several orbits, as well as times when radial migration appears to stall. Between $t = 15$ and 19 ORP, the pattern of outward migration for 2 ORP followed by an inward plunge over the next 2 ORP resembles the behavior in the $t = 0$ planet run during the burst, but, in this case, there is no distinct transition between phases of GI activity. Animations of the simulation (available at ???URL???) 
%***DICK: Fill in ???URL???
show there is a complex interaction between the planet, its spiral wake, and the global spiral arms caused by the instabilities. Analysis of images and periodicities suggests that, between 15 and 19 ORP, the planet may be in a mean motion resonance with $m = 2$ and 3 patterns, both of which have CR at 27 AU. Over this time interval, the average orbit period of the planet is $\sim$ 0.50 ORP, while both the $m = 2$ and 3 spirals have a pattern period of $\sim 0.74$ ORP, suggesting a 3:2 resonance. Both of these modes are present, at nearly the same periodicities, in the fiducial run. So, in contrast with the $t = 0$ planet run, this planet's rapid phases of radial migration are caused by interaction with strong GI modes that exist independent of the planet.

>>>>

\begin{abstract}
Characterization of migration in gravitationally unstable disks is necessary to understand the fate of protoplanets formed by disk instability. As part of a larger study, we are using a 3D radiative hydrodynamics code to investigate how an embedded gas giant planet interacts with a gas disk that undergoes gravitational instabilities (GIs). This Letter presents preliminary results from simulations with a Jupiter-mass planet placed in orbit at 25 AU within a 0.14 $M_{\odot}$ disk. The disk spans 5 to 40 AU around a 1 $M_{\odot}$ star and is initially close to marginal stability. In one simulation, the planet is inserted prior to the eruption of GIs; in another, it is inserted only after the disk has settled into a quasi-steady GI-active state, where heating by GIs roughly balances radiative cooling. When the planet is present from the beginning, its own wake stimulates growth of a particular global mode with which it strongly interacts, and the planet plunges inward six AU in about 10$^3$ years. In both cases with embedded planets, there are times when the planet's radial motion is slow and varies in direction. At other times, when the planet appears to be interacting with strong spiral modes, migration both inward and outward can be relatively rapid, covering several AUs over hundreds of years. Migration in both cases appears to stall near the inner Lindblad resonance of a dominant low-order mode. Planet orbit eccentricities fluctuate rapidly between about 0.02 to 0.1 throughout the GI-active phases of the simulations. 
\end{abstract}

@@

{\it The $t = 0$ Planet Run.} Figure \ref{fig:Am} shows that inclusion of even this modest mass planet ($0.7 \%$ of $M_d$) has a dramatic effect on the burst phase. Without a planet, when all modes grow from imposed noise, it takes 4 ORPs for coherent spiral modes to organize and grow to nonlinear amplitude. The growth is centered near the $Q$-minimum at $26$ AU. Modes of cos($m\phi$) symmetry with $m = 4$ and 5 dominate. An $m = 3$ (three-armed) spiral also grows, but somewhat more slowly, and lags the other modes in time. In the $t = 0$ planet run, the planet develops an organized wake within about an ORP in which $m = 3$ is a dominant component ($\sim 5$ \% global density perturbation). This nonlinear seeding causes $m = 3$ to dominate the GI burst, which also occurs $\sim$ 2 ORP earlier. Because of our initial placement of the planet inside but close to the $Q$-minimum, the corotation radius (CR) of the $m = 3$ mode is fairly close to that of the planet's orbit period. When the triggered $m = 3$ mode becomes strongly nonlinear at about 3 ORP, planet migration is significantly affected.

Figure \ref{fig:a} shows the evolution of the planet's radial position. From 0 to 2.5 ORP, the planet is torqued primarily by its own wake and migrates inward. 
Beginning at about 3 ORP, the planet interacts with the now nonlinear $m = 3$ GI mode. At first, the planet gains angular momentum and moves outward, but, from 4 to 8 ORP, a time interval of  about 720 yr, the plant experiences a negative torque and plunges from 23 to 17 AU.
After $t = 8$ ORP, the main burst is over, and the disk transitions into its asymptotic state, where modes of many $m$-values become comparably strong (see Fig. \ref{fig:Am}). The planet's radial migration apparently stalls at about 16 to 17 AU. From an analysis of periodicities present in the gas disk between 8 and 12 ORP, the planet lies a few AU inside the inner Lindblad resonance of a strong $m = 2$ mode with CR at 29 AU in this early part of the asymptotic phase.

{\it The $t = 10$ Planet Run.} The motion of the planet in this case is more difficult to interpret. We see intervals of fairly rapid inward or outward migration over several orbits, as well as times when radial migration appears to stall. Between $t = 15$ and 19 ORP, the pattern of outward migration for 2 ORP followed by an inward plunge over the next 2 ORP resembles the behavior in the $t = 0$ planet run during the burst, but, in this case, there is no distinct transition between phases of GI activity. Animations of the simulation (available at http://hdl.handle.net/2022/13304) show there is a complex interaction between the planet, its spiral wake, and the global spiral arms caused by the instabilities. Analysis of images and periodicities suggests that, between 15 and 19 ORP, the planet may be in a mean motion resonance with $m = 2$ and 3 patterns, both of which have CR at 27 AU. Over this time interval, the average orbit period of the planet is $\sim$ 0.50 ORP, while both the $m = 2$ and 3 spirals have a pattern period of $\sim 0.74$ ORP, suggesting a 3:2 resonance. Both of these modes are present, at nearly the same periodicities, in the fiducial run. So, in contrast with the $t = 0$ planet run, this planet's rapid phases of radial migration are caused by interaction with strong GI modes that exist independent of the planet.
